---
title: "The t-Test"
subtitle: "From Sampling Uncertainty to Statistical Inference"
author: "K.J.Mhango"
css: "scroll.css"
format:
  revealjs:
    theme: simple
    slide-number: true
    transition: slide
    incremental: false
execute:
  echo: true
  warning: false
  message: false
  freeze: auto
---

```{r}
#| label: setup
#| include: false
set.seed(123)
library(ggplot2)
library(dplyr)
library(tidyr)
```

# Welcome

## What this session is about

By the end of this session, you should be able to explain:

- why a sampling distribution matters
- what a z-score is doing
- why the t-statistic looks like a z-score
- what the t-distribution actually represents
- why the t-distribution depends on degrees of freedom
- how p-values and critical values come from areas under the curve

---

# Part 1 — The problem

## You mostly ever see one sample

In real research, you do not observe the whole population.

You observe one sample.

That sample gives you one sample mean:

$$
\bar{x}
$$

But your sample was only one possible sample.

If you had repeated the study, you would probably have obtained a slightly different mean.

---

## The first big idea

- A sample statistic is not fixed.
- It varies from sample to sample.
- So when we calculate a sample mean, we focus on this idea:  
  - Random sampling produces a range of possible sample means.
- That question leads to the idea of a **sampling distribution**.

---

## A sampling distribution is a distribution of statistics

Imagine repeating the same study many times.

Each time:

1. take a sample
2. calculate the sample mean
3. store that sample mean

After many repetitions, you would have a distribution of sample means.

That is the sampling distribution of $\bar{x}$.

---

## Simulating a sampling distribution

```{r}
#| label: simulate-sampling-distribution
population <- rnorm(100000, mean = 100, sd = 15)
sample_means <- replicate(
  5000,
  mean(sample(population, size = 25, replace = TRUE))
)
sampling_df <- data.frame(sample_mean = sample_means)
ggplot(sampling_df, aes(x = sample_mean)) +
  geom_histogram(aes(y = after_stat(density)), bins = 45) +
  geom_vline(xintercept = mean(population), linewidth = 1) +
  theme_classic() +
  labs(
    title = "A sampling distribution of sample means",
    x = "Sample mean",
    y = "Density"
  )
```

---

## The meaning of this plot

- The x-axis shows possible sample means.
- The y-axis shows probability density: where sample means tend to concentrate.
- The centre is close to the true population mean.
- The spread tells us how much the sample mean varies from sample to sample.

---

## Standard error

Sample means  have less variance than individual observation statistics.

If the population standard deviation is $\sigma$, then the standard deviation of the sample mean is:

$$
SE = \frac{\sigma}{\sqrt{n}}
$$

This is called the **standard error**. Technically we should use the sample standard deviation, but it is often approximated by the population standard deviation.

Remember: Standard in standard error just means the numerator is normalised by

It measures the expected sampling variability of $\bar{x}$.

---

# Part 2 — Standardising makes differences interpretable

## The z-statistic

If the population standard deviation is known, we can calculate:

$$
Z = \frac{\bar{x} - \mu}{\sigma / \sqrt{n}}
$$

This tells us how many standard errors the sample mean is away from $\mu$.

---

## The hidden assumption

- The z-statistic assumes that $\sigma$ is known.
- That is a very strong assumption.
- In many real experiments, especially small experiments, we do not know the population standard deviation.

We estimate it from the same sample:

$$
s
$$

This changes everything.

---

# Part 3 — From z to t

## The t-statistic

When $\sigma$ is unknown, we replace it with $s$:

$$
t = \frac{\bar{x} - \mu}{s / \sqrt{n}}
$$

This looks almost identical to a z-statistic but it is not the same object.

The denominator now contains an estimate from the data.

---

## You only calculate one t-statistic

In practice, you do not calculate a whole distribution.

You calculate one number:

$$
t = \frac{\bar{x} - \mu}{s / \sqrt{n}}
$$

Then you evaluate this claim:

> This value may be unusual relative to what we would expect if the null hypothesis were true.

To answer that, you need a reference distribution.

That reference distribution is the **t-distribution**.

---

# Part 4 — The t-distribution as a reference distribution

## The t-distribution as a thought experiment

The t-distribution is the distribution of t-statistics you would get if you repeated the same experiment many times under the null hypothesis.

Each repetition would produce a new sample.

Each sample would produce:

- a sample mean $\bar{x}$
- a sample standard deviation $s$
- a t-statistic

The distribution of those t-statistics is the t-distribution.

---

## Simulating the t-distribution

```{r}
#| label: simulate-t-function
simulate_t <- function(n, reps = 10000) {
  t_values <- numeric(reps)
  for (i in seq_len(reps)) {
    x <- rnorm(n, mean = 0, sd = 1)
    xbar <- mean(x)
    s <- sd(x)
    t_values[i] <- (xbar - 0) / (s / sqrt(n))
  }
  data.frame(t = t_values)
}
t_df_5 <- simulate_t(n = 5)
```

---

## A simulated t-distribution

```{r}
#| label: plot-simulated-t
n <- 5
df <- n - 1
ggplot(t_df_5, aes(x = t)) +
  geom_histogram(aes(y = after_stat(density)), bins = 60) +
  stat_function(fun = dt, args = list(df = df), linewidth = 1) +
  theme_classic() +
  labs(
    title = "A simulated t-distribution",
    subtitle = paste0("n = ", n, ", df = ", df),
    x = "Possible t-statistic values",
    y = "Density"
  )
```

---

## The x-axis

The x-axis contains possible values of the t-statistic.

For example:


-3, -2, -1, 0, 1, 2, 3


- A value near 0 means the sample mean is close to the null value.
- A large positive value means the sample mean is far above the null value.
- A large negative value means the sample mean is far below the null value.

---

## The y-axis

- The y-axis is **probability density**.
- It is not probability by itself.
- Probability comes from **area under the curve**.
- The total area under the whole t-distribution is 1.
- That represents all possible t-statistic values.

---

## Where your t-statistic goes

- After calculating your t-statistic, you place it on the x-axis.
- If it lands near the centre, it is ordinary under the null hypothesis.
- If it lands far into a tail, it is unusual under the null hypothesis.
- The test asks how much area lies beyond your observed value.

---

# Part 5 — The t-distribution depends on df

## Degrees of freedom

For a one-sample t-test:

$$
df = n - 1
$$

The degrees of freedom are linked to how much independent information remains for estimating the sample variance.

Once the sample mean is estimated, the deviations from the mean are constrained to sum to zero.

That leaves $n - 1$ freely varying deviations.

---

## Why df changes the shape

- The t-statistic uses $s$, not $\sigma$.
- When $n$ is small, $s$ is a noisy estimate of the true standard deviation.
- Sometimes $s$ is too large.
- Sometimes $s$ is too small.
- When $s$ is too small, the t-statistic can become very large.
- That makes extreme t-values more common.

---

## Heavy tails

- Small degrees of freedom produce heavier tails.
- This means the distribution expects more extreme t-statistics.
- It is more cautious because the denominator is uncertain.
- As degrees of freedom increase, $s$ becomes more stable.
- The t-distribution gradually approaches the normal distribution.

---

## Comparing different degrees of freedom

```{r}
#| label: compare-t-distributions
x_vals <- seq(-5, 5, length.out = 500)
compare_df <- data.frame(
  x = x_vals,
  `df = 2` = dt(x_vals, df = 2),
  `df = 5` = dt(x_vals, df = 5),
  `df = 30` = dt(x_vals, df = 30),
  `Normal` = dnorm(x_vals)
) |>
  pivot_longer(-x, names_to = "distribution", values_to = "density")
ggplot(compare_df, aes(x = x, y = density, colour = distribution)) +
  geom_line(linewidth = 1) +
  theme_classic() +
  labs(
    title = "The t-distribution changes with degrees of freedom",
    x = "Possible t-statistic values",
    y = "Density",
    colour = "Distribution"
  )
```

---

## The important interpretation

- A t-value of 2 does not always mean the same thing.
- It depends on the degrees of freedom.
- With low df, a t-value of 2 is less surprising because the distribution is wider.
- With high df, a t-value of 2 is more surprising because the distribution is tighter.
- That is why the t-test must know the df.

---

# Part 6 — Using area under the t-distribution

## Probability comes from area

- The curve gives density.
- The area under the curve gives probability.
- Once we calculate a t-statistic, we make this area-based interpretation:
    - The area beyond this value determines how unusual it is under the null model.
- That area is the basis of the p-value.

---

## Right-tail area

```{r}
#| label: right-tail-area
observed_t <- 1.86
df <- 9
right_tail <- 1 - pt(observed_t, df = df)
right_tail
```

This is the probability of getting a t-statistic at least this large under the null hypothesis.

---

## Left-tail area

```{r}
#| label: left-tail-area
left_tail <- pt(-observed_t, df = df)
left_tail
```

This is the probability of getting a t-statistic this small or smaller under the null hypothesis.

---

## Two-tailed area

```{r}
#| label: two-tailed-area
two_tailed <- 2 * (1 - pt(abs(observed_t), df = df))
two_tailed
```

This is the probability of getting a t-statistic at least this extreme in either direction.

This is the usual p-value for a two-tailed t-test.

---

## Visualising the observed t-statistic

```{r}
#| label: observed-t-on-distribution
plot_df <- data.frame(x = seq(-4, 4, length.out = 500)) |>
  mutate(density = dt(x, df = df))
ggplot(plot_df, aes(x = x, y = density)) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = observed_t, linewidth = 1) +
  theme_classic() +
  labs(
    title = "Place your observed t-statistic on the t-distribution",
    subtitle = paste0("Observed t = ", observed_t, ", df = ", df),
    x = "t-statistic",
    y = "Density"
  )
```

---

# Part 7 — Critical values

## Going backwards from area to cutoff

Sometimes we choose the area first.

For example, in a two-tailed test with:

$$
\alpha = 0.05
$$

we place 2.5% of the area in the left tail and 2.5% in the right tail.

The t-values that mark those tail boundaries are called **critical values**.

---

## Calculating a two-tailed critical value

```{r}
#| label: two-tailed-critical
alpha <- 0.05
df <- 9
t_critical <- qt(1 - alpha / 2, df = df)
t_critical
```

For $df = 9$, the two-tailed 5% critical values are:

$$
-t^* \quad \text{and} \quad +t^*
$$

---

## Visualising critical values

```{r}
#| label: plot-critical-values
ggplot(plot_df, aes(x = x, y = density)) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = c(-t_critical, t_critical), linewidth = 1) +
  geom_vline(xintercept = observed_t, linetype = "dashed", linewidth = 1) +
  theme_classic() +
  labs(
    title = "Critical values create rejection regions",
    subtitle = paste0("Two-tailed alpha = ", alpha, ", df = ", df),
    x = "t-statistic",
    y = "Density"
  )
```

---

## How to interpret the critical values

If the observed t-statistic is beyond the critical value, it falls into a region we agreed to treat as unusually extreme.

For a two-tailed test:

$$
|t| > t^*
$$

means reject the null hypothesis.

If the observed t-statistic is inside the critical boundaries, we do not reject the null hypothesis.

---

## One-tailed critical values

For a right-tailed test with $\alpha = 0.05$, all 5% of the rejection area is placed in the right tail.

```{r}
#| label: one-tailed-critical
qt(0.95, df = 9)
```

This cutoff is smaller than the two-tailed cutoff because the rejection area is concentrated in one tail.

---

# Part 8 — Manual worked example

## Suppose we have these values

$$
\bar{x} = 101.3
$$

$$
\mu_0 = 100
$$

$$
s = 2.2
$$

$$
n = 10
$$

We test whether the true mean differs from 100.

---

## Calculate the t-statistic

```{r}
#| label: manual-t-statistic
xbar <- 101.3
mu0 <- 100
s <- 2.2
n <- 10
t_stat <- (xbar - mu0) / (s / sqrt(n))
t_stat
```

---

## Compare with the critical value

```{r}
#| label: manual-critical-comparison
df <- n - 1
t_critical <- qt(0.975, df = df)
abs(t_stat) > t_critical
```

The observed t-statistic is not beyond the two-tailed critical value.

So we do not reject the null hypothesis at $\alpha = 0.05$.

---

## Calculate the p-value

```{r}
#| label: manual-p-value
p_value <- 2 * (1 - pt(abs(t_stat), df = df))
p_value
```

This p-value is the two-tailed area beyond the observed t-statistic.

---

# Part 9 — The three core t-tests

## One-sample t-test

The one-sample t-test compares one sample mean to a reference value.

You use it when:

- you have one group of observations
- you want to test against a known or hypothesised population mean $\mu_0$
- population standard deviation is unknown and estimated by $s$

The test statistic is:

$$
t = \frac{\bar{x} - \mu_0}{s / \sqrt{n}}
$$

Interpretation:

- numerator = observed difference from the null value
- denominator = estimated standard error of the sample mean
- large absolute t-values indicate stronger disagreement with the null model

---

## Two-sample t-test (independent groups)

The two-sample t-test compares the means of two independent groups.

You use it when:

- each observation belongs to only one group
- groups represent different people, units, or experimental conditions
- you want to test whether $\mu_1 - \mu_2 = 0$ (or another specified difference)

A common formulation is:

$$
t = \frac{\bar{x}_1 - \bar{x}_2}{\sqrt{\frac{s_1^2}{n_1} + \frac{s_2^2}{n_2}}}
$$

This is the usual form when the null difference is 0.

Interpretation:

- the numerator is the observed mean gap
- the denominator is the estimated uncertainty in that gap
- inference depends on the standard error model (equal-variance or unequal-variance)

---

## Paired t-test (dependent observations)

The paired t-test is a one-sample t-test on within-pair differences.

You use it when:

- each subject is measured twice (before/after)
- observations are naturally matched pairs (twins, matched controls, repeated measures)
- dependence exists within pairs and must be preserved in the analysis

Define the pairwise difference:

$$
d_i = x_{i,\text{before}} - x_{i,\text{after}}
$$

Then test:

$$
t = \frac{\bar{d}}{s_d / \sqrt{n}}
$$



Interpretation:

- the unit of analysis is the difference, not each raw measurement separately
- pairing often reduces noise by removing stable person-level variability
- smaller noise in differences can increase power

---

## Choosing among one-sample, two-sample, and paired designs

Use this quick logic:

- one-sample: one group compared with a benchmark
- two-sample: two independent groups compared with each other
- paired: two measurements per unit, analysed via differences

The design determines the correct t-test.

Choosing the wrong design usually gives the wrong standard error and invalid inference.

---

# Part 10 — Assumptions behind the t-test family

## Core assumptions

All t-tests rely on structural assumptions:

- data values (or pairwise differences) are sampled independently
- measurements are on a continuous or near-continuous scale
- extreme outliers do not dominate mean and standard deviation estimates

When sample sizes are moderate or large, t-tests are often robust to mild non-normality.

---

## Distributional assumption

For small samples, the relevant analysis unit should be approximately normal:

- one-sample t-test: raw values in the single group
- paired t-test: within-pair differences
- two-sample t-test: each group distribution should be reasonably well-behaved

Perfect normality is not required, but severe skewness plus very small $n$ weakens reliability.

---

## Equal-variance assumption for the pooled two-sample test

The classical pooled two-sample t-test assumes:

$$
\sigma_1^2 = \sigma_2^2
$$

If this is badly violated, pooled standard errors and p-values can be misleading.

That is why the unequal-variance version is often preferred in practice.

---

## Practical diagnostics before running a t-test

Useful checks include:

- inspect boxplots or histograms for strong skew and outliers
- verify study design supports independence
- confirm whether data are paired or independent before choosing the test
- compare group spreads in two-sample settings

Diagnostics guide model choice; they do not replace domain judgement.

---

# Part 11 — Unequal variances: Welch's t-test

## Why Welch's test exists

When two groups have different variances, forcing a pooled variance can distort inference.

Welch's t-test avoids this by estimating the standard error without assuming equal variances.

It is the default for `t.test(x, y)` in R.

---

## Welch test statistic

Welch uses:

$$
t = \frac{(\bar{x}_1 - \bar{x}_2) - \Delta_0}{\sqrt{\frac{s_1^2}{n_1} + \frac{s_2^2}{n_2}}}
$$

with an approximate degrees-of-freedom formula (Welch-Satterthwaite).

The df can be non-integer, and that is expected.

---

## Pooled versus Welch in practice

Pooled two-sample t-test is appropriate only when equal variances are genuinely plausible.

Welch's test is usually safer because:

- it remains valid when variances differ
- it performs similarly when variances are actually equal
- it better handles unequal sample sizes combined with unequal variances

For most applied work, Welch should be the default independent two-sample t-test.

---

## R syntax for both versions

```{r}
# Welch two-sample t-test (default)
setosa_demo <- iris$Petal.Length[iris$Species == "setosa"]
versicolor_demo <- iris$Petal.Length[iris$Species == "versicolor"]
t.test(setosa_demo, versicolor_demo)

# Pooled two-sample t-test (equal variances forced)
t.test(setosa_demo, versicolor_demo, var.equal = TRUE)
```

The key difference is `var.equal = TRUE`.

If you do not set it, R performs Welch's test.

---

# Part 12 — Lab work

## Exercise 1 - the one sample t-test

You are a researcher studying the effect of a new drug on blood pressure.

You have a sample of 10 patients with a mean blood pressure of 120 mmHg.

The sample standard deviation is 10 mmHg.

Test whether the true mean blood pressure is different from 110 mmHg.

---

Let's perform the hypothesis test step by step.

- **Null hypothesis ($H_0$):** The true mean blood pressure is 110 mmHg ($\mu = 110$).
- **Alternative hypothesis ($H_A$):** The true mean is not 110 mmHg ($\mu \neq 110$).

Given:
- $n = 10$
- $\bar{x} = 120$
- $s = 10$
- $\mu_0 = 110$

### 1. Calculate the t-statistic

```{r}
xbar <- 120
mu0 <- 110
s <- 10
n <- 10
t_stat <- (xbar - mu0) / (s / sqrt(n))
t_stat
```

### 2. Degrees of freedom and critical value

```{r}
df <- n - 1
t_critical <- qt(0.975, df = df)
abs(t_stat) > t_critical
```

### 3. Calculate the p-value

```{r}
p_value <- 2 * (1 - pt(abs(t_stat), df = df))
p_value
```

### 4. Interpretation

If the absolute value of the t-statistic is greater than the critical value, or if the p-value < 0.05, we reject the null hypothesis and conclude that the drug has a statistically significant effect on blood pressure.

## Exercise 2 - the two sample t-test 

Use the `iris` dataset to compare the mean petal length between the species *setosa* and *versicolor*. Test whether there is a significant difference between the two means.

```{r}
#| label: iris-two-sample-ttest
# Subset Petal.Length for setosa and versicolor
setosa_petal <- iris$Petal.Length[iris$Species == "setosa"]
versicolor_petal <- iris$Petal.Length[iris$Species == "versicolor"]

# Perform a two-sample t-test (assume equal variance for now)
t_test_result <- t.test(setosa_petal, versicolor_petal, var.equal = TRUE)
t_test_result
```

The above output summarizes the two-sample t-test comparing Petal Length between *setosa* and *versicolor*.

- **Null hypothesis:** The population means of Petal Length for setosa and versicolor are equal.
- **Alternative hypothesis:** The population means are different.

Interpret the p-value:
- If the p-value < 0.05, there is a significant difference between the means.
- If the p-value ≥ 0.05, there is no significant difference.

## Exercise 3 - the paired t-test
## Exercise 3 - the paired t-test

A nutritionist wants to test whether a new dietary supplement has an effect on cholesterol levels. She measures the cholesterol levels of 8 patients before and after using the supplement for 2 months.

The results (in mg/dL) are as follows:

| Patient | Before | After |
|---------|--------|-------|
| 1       | 210    | 200   |
| 2       | 225    | 218   |
| 3       | 240    | 230   |
| 4       | 199    | 192   |
| 5       | 220    | 210   |
| 6       | 215    | 203   |
| 7       | 230    | 220   |
| 8       | 205    | 202   |

Test whether the supplement has a significant effect on mean cholesterol levels.

### 1. Enter the data and perform a paired t-test

```{r}
# Before and after cholesterol levels
before <- c(210, 225, 240, 199, 220, 215, 230, 205)
after <- c(200, 218, 230, 192, 210, 203, 220, 202)

# Perform paired t-test
paired_ttest <- t.test(before, after, paired = TRUE)
paired_ttest
```

### 2. Interpretation

- **Null hypothesis ($H_0$):** The mean difference in cholesterol levels (before - after) is zero (the supplement has no effect).
- **Alternative hypothesis ($H_A$):** The mean difference is not zero (the supplement has an effect).

Interpret the p-value from the output:

- If the p-value < 0.05, conclude that the supplement has a statistically significant effect on cholesterol levels.
- If the p-value ≥ 0.05, there is no significant evidence of an effect.

